---
title: "Comprehensive Introduction to R Programming for Data Science"
author: "Zorzi Enrico"
output:
  html_document:
    toc: true
    toc_float: true
    theme: default
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

# R Basics: Vectors, Matrices, and Data Frames

R is a versatile language for statistical computing and data analysis. Its fundamental data structures form the foundation of most data operations.

## Vectors

Vectors, the most basic data structure in R, are one-dimensional arrays that store elements of the same type (numeric, character, or logical).

```{r vectors}
# Creating vectors
numeric_vector <- c(1, 2, 3, 4, 5)
character_vector <- c("apple", "banana", "cherry")
logical_vector <- c(TRUE, FALSE, TRUE)

# Vector operations
sum_vector <- numeric_vector + 10
mean_value <- mean(numeric_vector)

# Filtering values greater than 3
filtered_vector <- numeric_vector[numeric_vector > 3]
```

Vectors are created using the `c()` function. Basic operations include element-wise addition and summary statistics like the mean. Logical vectors are crucial for filtering data.

## Matrices

Matrices are two-dimensional arrays holding data of the same type, suitable for tabular numeric data. They are widely used in linear algebra and machine learning.

```{r matrices}
# Creating a matrix
matrix_data <- matrix(1:9, nrow = 3, ncol = 3)

# Matrix operations
transposed_matrix <- t(matrix_data)
matrix_product <- matrix_data %*% transposed_matrix
```

Real-world application: Matrices are used in neural networks for forward and backward propagation.

## Data Frames

Data frames, the most commonly used data structure in R, allow storage of mixed data types. They are ideal for working with datasets in data science.

```{r dataframes}
# Creating a data frame
df <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),
  Age = c(25, 30, 35),
  Height = c(165, 180, 175)
)

# Accessing data frame elements
print(df$Name)  # Accessing a column
print(df[1, 2])  # Accessing a specific element
```

Real-world application: Data frames are used in machine learning for handling datasets during preprocessing.

# Data Visualization: ggplot2

Data visualization is crucial for understanding and communicating insights. The `ggplot2` package is part of the tidyverse ecosystem and provides a powerful system for creating high-quality plots.

## Basic Plots

```{r basic_plots}
# Basic scatter plot
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  labs(title = "Car Weight vs. MPG",
       x = "Weight (1000 lbs)",
       y = "Miles per Gallon")

# Bar plot with customization
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  theme_minimal() +
  labs(title = "Average MPG by Cylinder Count",
       x = "Number of Cylinders",
       y = "Miles per Gallon")
```

## Customization and Themes

`ggplot2` allows for extensive customization through themes and annotations.

```{r customized_plots}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = "blue", size = 3) +
  theme_classic() +
  labs(title = "Customized Scatter Plot",
       subtitle = "Demonstrating ggplot2's flexibility")
```

# Statistical Analysis: t-tests and ANOVA

## t-tests

`t.test()` in R is used for comparing means. It is essential for hypothesis testing.

```{r ttest}
# Example: Comparing MPG for two transmission types
t_result <- t.test(mpg ~ am, data = mtcars)
print(t_result)
```

Real-world application: A/B testing in product development.

## ANOVA

ANOVA is used to test differences between multiple groups.

```{r anova}
# Example: ANOVA for MPG across cylinder groups
anova_result <- aov(mpg ~ factor(cyl), data = mtcars)
summary(anova_result)
```

Applications include experimental designs and clinical trials.

# Advanced Topics: Bioinformatics with Bioconductor

`Bioconductor` provides specialized tools for bioinformatics.

## Gene Expression Analysis

Gene expression data can be analyzed for differential expression.

```{r gene_expression, eval=FALSE}
library(DESeq2)

# Example data loading and DESeq analysis
dds <- DESeqDataSetFromMatrix(countData = count_matrix,
                             colData = col_data,
                             design = ~ condition)
dds <- DESeq(dds)
results <- results(dds)
```

## Heatmaps

Heatmaps visualize complex data matrices.

```{r heatmap, eval=FALSE}
library(pheatmap)

pheatmap(matrix_data,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         main = "Heatmap Example")
```

# Conclusion

R programming is a powerful tool for data science, offering comprehensive functionality for data manipulation, visualization, and statistical analysis. Mastering its basics and advanced features can significantly enhance data-driven decision-making.
